---
title: "Palmer Penguins"
author: "Dr.Wonder Unaji"
date: "2023-10-05"
output:html_document
---
## Loading the libraries used both 'ggplot' and 'palmerpenguins'
```{r}
library(ggplot2)
library(palmerpenguins)
```
Here all i did was load the packages to be worked with

## Working with the data
```{r penguins,echo=FALSE}
ggplot(data=penguins, aes(x=flipper_length_mm, y=body_mass_g)) + 
  geom_point() 
```
Having loaded the packages earlier in the last code chunk.I used the ggplot function then then the geom_point function to generate the scatter plot using the the two relationships which are flipper_length_mm and body_mass_g 

## Modifying the visual
```{r}
ggplot(data=penguins, aes(x=flipper_length_mm, y=body_mass_g,colour =sex,shape =species)) + 
  geom_point() 
```
Modifying the data for easy visibility by introducing colour and shape variables.

## Dividing the relationship based on their species
````{r}
ggplot(data=penguins, aes(x=flipper_length_mm, y=body_mass_g,colour =sex,shape =species))+
  geom_point()+
  facet_wrap(~species)
````
Here i used the facet function to show a subset of the entire plot using their species i.e Adelies,chinstrap and gentoo so we could study the visual to see the trends individually.

## Adding a Title and Subtitle 
````{r}
ggplot(data=penguins, aes(x=flipper_length_mm, y=body_mass_g,colour =sex,shape =species))+
  geom_point()+
  facet_wrap(~species)+
  labs(subtitle="Palmer Penguins:Flipper_length_mm vs Body_mass_g using faceting",caption = "Data collected by Dr Elon Church")
````
Improved visualisations by adding title and source of the data using the labs function

## Assigning a variable to represent the enitre code for easy process
````{r}
W <- ggplot(data=penguins, aes(x=flipper_length_mm, y=body_mass_g,colour =sex,shape =species))+
  geom_point()+
  facet_wrap(~species)+
  labs(subtitle="Palmer Penguins:Flipper_length_mm vs Body_mass_g using faceting",caption = "Data collected by Dr Elon Church")
````
For easy referencing as we proceed I assigned all the data in the code chunk to the variable W
## Formatting texts Final
````{r}
W+ theme(
  plot.subtitle = element_text(family = "Arial", size = 16, face = "bold"),
  axis.title.x = element_text(family = "Arial", size = 12),
  axis.title.y = element_text(family = "Arial", size = 12,)
)
````
Formatting and final adjustments to data


